A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-05-30 sections/Metatheory.tex
shrinking text

To simplify matters, we introduce an intermediate system that is still declarative
in nature, but has no overlap. This intermediate system differs only in a few 
rules from DK's declarative system; these are the changes:
\begin{enumerate}
\item Restrict the shape of $B$ in the rule $\mathtt{\forall L}$ subtyping rule:
$$
\inferrule*[right=$\mathtt{\forall L'}$]
{B \neq \all[b]B' \\ \Psi \vdash \tau \\ \Psi \vdash [\tau/a]A \le B}
{\Psi \vdash \all A \le B}
$$
\item Drop the redundant rule $\mathtt{Decl1I}$,
which can be easily derived by a combination of
$\mathtt{DeclSub}$, $\mathtt{Decl1I{\To}}$ and $\mathtt{{\le}Unit}$:
$$
\inferrule*[right=$\mathtt{DeclSub}$]
{
    \inferrule*[right=$\mathtt{Decl1I{\To}}$]
    {~}
    {\Psi \vdash () \To 1}
    \\
    \inferrule*[right=$\mathtt{{\le}Unit}$]
    {~}
    {\Psi \vdash 1 \le 1}
}
{\Psi \vdash () \Lto 1}
$$
\item Restrict the shapes of $e$ and $A$ in the subsumption rule $\mathtt{DeclSub}$:
$$
\inferrule*[right=$\mathtt{DeclSub'}$]
{e \neq \lam e' \\ A \neq \all A' \\ \Psi \vdash e \To A \\ \Psi \vdash A \le B}
{\Psi \vdash e \Lto B}
$$
\end{enumerate}
The resulting declarative system has no overlapping rules
and more closely resembles the algorithmic system,
which contains constraints of the same shape.

@@

Our algorithm is sound with respect to DK's declarative system.
For any worklist $\Gm$ that reduces successfully,
there is a valid instantiation $\Om$ that transfers all judgments successfully to the declarative system.
\begin{theorem}[Soundness]
If \emph{wf }$\Gm$ and $\Gm \redto \nil$,
then there exists $\Om$ s.t. $\Gm\sto\Om$ and $\Om\redto\nil$.
\end{theorem}

@@

The completeness of our algorithm means that any derivation in the
declarative system has an algorithmic counterpart.

@@

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        \\then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        \\then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \Vdash e \To_a 1 \le 1 \sto \Om \Vdash e \To_a 1 \le 1$
        \\then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \le 1]$
        \\then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

Finally, we show that our algorithm is decidable:
\begin{theorem}[Decidability]
Given a well-formed input $\Gm$, either $\Gm\redto\nil$ or not.
\end{theorem}
Our decidability proof is based on a lexicographic group of induction measures\\
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
on the worklist $\Gm$. Formal definitions of these measures can be found in
Figure~\ref{fig:measures} in the Appendix. The first two, 
$|\cdot|_e$ and $|\cdot|_\Leftrightarrow$, measure the total size of terms
and the total difficulty of judgments, respectively. In the latter, check judgments
count for 2, inference judgments for 1 and function inference judgments for 3.
Another two measures, $|\cdot|_\forall$ and $|\cdot|_\to$, count the total number of
universal quantifiers and function types, respectively. Finally,
$|\cdot|_{\al}$ counts the number of existential variables in the worklist,
and $|\cdot|$ is simply the length of the worklist.

@@

We now show that entirely reducing these smaller problems leaves the worklist
in a state with an overall smaller measure. Our starting point is a 
worklist of the form $\Gm,\Gm_i$ where $\Gm_i$ are instantiation judgments.
\begin{gather*}
\begin{aligned}
\Gm_i &:= \nil \mid \Gm_i, \al\le A \mid \Gm_i, A\le\al \quad
    \text{where } \al\notin FV(A) \cup FV(\Gm_i)
\end{aligned}
\end{gather*}
Fully reducing these instantiation judgments at the top of the worklist
has a twofold impact. Firstly, new entries may be pushed onto the worklist which
are not instantiation judgments. This only happens when $\Gm_i$ contains a universal quantifier
that is reduced by Rule 8 or 9. The new entries then are of the form $\Gm_\le$:
\begin{gather*}
\begin{aligned}
\Gm_\le &:= \nil \mid \Gm_\le, a \mid \Gm_\le, \al \mid \Gm_\le, A\le B
\end{aligned}
\end{gather*}
Secondly, reducing the instantiation judgments may also affect the remainder of the worklist $\Gm$,
by solving existing existentials and introducing new ones. This worklist update is captured in the 
update judgment $\Gm \jExt \Gm'$ defined in Figure~\ref{fig:worklist_ext}.
For instance, an existential variable instantiation, 
$\Gm_L,\al,\Gm_R \jExt \Gm_L, \al[1], \al[2], [\al[1] \to \al[2] / \al]\Gm_R$,
can be derived as a combination of the three rules that define the update relation.

@@

We have chosen the Abella (v2.0.7-dev
\footnote{We use a development version because the developers
just fixed a serious bug that accepts a simple proof of false,
which also affects the completeness of one of our proofs.
Specifically, our scripts compile against commit 92829a of Abella's GitHub repository.}
) proof assistant~\cite{AbellaDesc} to develop our formalization.
Abella is designed to help with formalizations of programming languages,
due to its built-in support for variable binding and
the $\lambda$-tree syntax~\cite{miller2000abstract}, which is a form of HOAS.
Nominal variables, or $\nabla$-quantified variables, are used as an unlimited name supply,
which supports explicit freshness control and substitutions.
Although Abella lacks packages, tactics and support for modules,
its higher-order unification and the ease of formalizing substitution-intensive relations
are very helpful for our formalization.

While the algorithmic rules are in a small-step style,
the proof script rewrites them into a big-step style,
where inductions are easier to handle.
In addition, we do prove the equivalence of the two representations.

\paragraph{Statistics of the Proof}
The proof script consists of 7977 lines of Abella code with a total of
51 definitions and 596 theorems.
Figure~\ref{fig:proof_statistics} briefly summarizes the contents of each file.
The files are linearly dependent due to limitations of Abella.

\begin{figure}[ht]
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  311 & 57  & Basic data structures\\\hline
    typing.thm          &  245 & 7   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strengthening\\\hline
    alg.thm             &  679 & 80  & Basic algorithmic properties\\\hline
    trans.thm           &  616 & 53  & \makecell[l]{Worklist instantiation and declarative transfer\\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  909 & 70  & \makecell[l]{Non-overlapping declarative system \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll},
                                \ref{lem:inv_chkLam}, \ref{lem:subsumption}}\\\hline
    soundness.thm       & 1107 & 78  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & The $|\cdot|_\to$ measure; Lemma~\ref{lem:prune_inst}\\\hline
    dcl.thm             &  380 & 12  & Non-overlapping declarative worklist \\\hline
    completeness.thm    & 1124 & 61  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  837 & 45  & Other worklist measures; Lemma~\ref{lemma:inst:decidable}\\\hline
    decidability.thm    &  983 & 57  & Decidability theorem and corollary\\\hline
    smallStep.thm       &  119 &  2  & The equivalence between big-step and small-step\\\hline
    \hline\emph{Total}               & 7977 & 596 & (51 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}

>>>>

To simplify matters, we introduce an intermediate system that is still declarative
in nature, but has no overlap. This intermediate system differs only in a few 
rules from DK's declarative system:
\begin{enumerate}
\item Restrict the shape of $B$ in the rule $\mathtt{\forall L}$ subtyping rule:
$$
\inferrule*[right=$\mathtt{\forall L'}$]
{B \neq \all[b]B' \\ \Psi \vdash \tau \\ \Psi \vdash [\tau/a]A \le B}
{\Psi \vdash \all A \le B}
$$
\item Drop the redundant rule $\mathtt{Decl1I}$,
which can be easily derived by a combination of
$\mathtt{DeclSub}$, $\mathtt{Decl1I{\To}}$ and $\mathtt{{\le}Unit}$:
$$
\inferrule*[right=$\mathtt{DeclSub}$]
{
    \inferrule*[right=$\mathtt{Decl1I{\To}}$]
    {~}
    {\Psi \vdash () \To 1}
    \\
    \inferrule*[right=$\mathtt{{\le}Unit}$]
    {~}
    {\Psi \vdash 1 \le 1}
}
{\Psi \vdash () \Lto 1}
$$
\item Restrict the shapes of $e$ and $A$ in the subsumption rule $\mathtt{DeclSub}$:
$$
\inferrule*[right=$\mathtt{DeclSub'}$]
{e \neq \lam e' \\ A \neq \all A' \\ \Psi \vdash e \To A \\ \Psi \vdash A \le B}
{\Psi \vdash e \Lto B}
$$
\end{enumerate}
The resulting declarative system has no overlapping rules
and more closely resembles the algorithmic system,
which contains constraints of the same shape.

@@

Our algorithm is sound with respect to DK's declarative system.
For any worklist $\Gm$ that reduces successfully,
there is a valid instantiation $\Om$ that transfers all judgments
to the declarative system.
\begin{theorem}[Soundness]
If \emph{wf }$\Gm$ and $\Gm \redto \nil$,
then there exists $\Om$ s.t. $\Gm\sto\Om$ and $\Om\redto\nil$.
\end{theorem}

@@

The completeness of our algorithm means that any derivation in the
declarative system has an algorithmic counterpart.
% \jimmy{In response to "treat declarative and algorithmic worklists as a subsets":}
We explicitly relate between an algorithmic context $\Gm$ and
a declarative context $\Omega$ to avoid potential confusion.

@@

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgments:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        \\then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        \\then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \Vdash e \To_a 1 \le 1 \sto \Om \Vdash e \To_a 1 \le 1$
        \\then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \le 1]$
        \\then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

Finally, we show that our algorithm is decidable:
\begin{theorem}[Decidability]
Given \emph{wf} $\Gm$, it is decidable whether $\Gm\redto\nil$ or not.
\end{theorem}
Our decidability proof is based on a lexicographic group of induction measures\\
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
on the worklist $\Gm$. Formal definitions of these measures can be found in
Figure~\ref{fig:measures} in the Appendix. The first two, 
$|\cdot|_e$ and $|\cdot|_\Leftrightarrow$, measure the total size of terms
and the total difficulty of judgments, respectively. In the latter, check judgments
count for 2, inference judgments for 1 and function inference judgments for 3.
Another two measures, $|\cdot|_\forall$ and $|\cdot|_\to$, count the total number of
universal quantifiers and function types, respectively. Finally,
$|\cdot|_{\al}$ counts the number of existential variables in the worklist,
and $|\cdot|$ is simply the length of the worklist.

@@

We now show that entirely reducing these smaller problems leaves the worklist
in a state with an overall smaller measure. Our starting point is a 
worklist $\Gm,\Gm_i$ where $\Gm_i$ are instantiation judgments.
\begin{gather*}
\begin{aligned}
\Gm_i &:= \nil \mid \Gm_i, \al\le A \mid \Gm_i, A\le\al \quad
    \text{where } \al\notin FV(A) \cup FV(\Gm_i)
\end{aligned}
\end{gather*}
Fully reducing these instantiation judgments at the top of the worklist
has a twofold impact. Firstly, new entries may be pushed onto the worklist which
are not instantiation judgments. This only happens when $\Gm_i$ contains a universal quantifier
that is reduced by Rule 8 or 9. The new entries then are of the form $\Gm_\le$:
\begin{gather*}
\begin{aligned}
\Gm_\le &:= \nil \mid \Gm_\le, a \mid \Gm_\le, \al \mid \Gm_\le, A\le B
\end{aligned}
\end{gather*}
Secondly, reducing the instantiation judgments may also affect the remainder of the worklist $\Gm$,
by solving existing existentials and introducing new ones. This worklist update is captured in the 
update judgment $\Gm \jExt \Gm'$ defined in Figure~\ref{fig:worklist_ext}.
For instance, an existential variable instantiation, 
$\Gm_L,\al,\Gm_R \jExt \Gm_L, \al[1], \al[2], [\al[1] \to \al[2] / \al]\Gm_R$,
can be derived as a combination of the three rules that define the update relation.

@@

We have chosen the Abella (v2.0.7-dev
\footnote{We use a development version because the developers
just fixed a serious bug that accepts a simple proof of false,
which also affects our proof.
Specifically, our scripts compile against commit 92829a of Abella's GitHub repository.}
) proof assistant~\cite{AbellaDesc} to develop our formalization.
Abella is designed to help with formalizations of programming languages,
due to its built-in support for variable binding and
the $\lambda$-tree syntax~\cite{miller2000abstract}, which is a form of HOAS.
Nominal variables, or $\nabla$-quantified variables, are used as an unlimited name supply,
which supports explicit freshness control and substitutions.
Although Abella lacks packages, tactics and support for modules,
its higher-order unification and the ease of formalizing substitution-intensive relations
are very helpful.

While the algorithmic rules are in a small-step style,
the proof script rewrites them into a big-step style
for easier inductions.
In addition, we do prove the equivalence of the two representations.

\paragraph{Statistics of the Proof}
The proof script consists of 7977 lines of Abella code with a total of
60 definitions and 596 theorems.
Figure~\ref{fig:proof_statistics} briefly summarizes the contents of each file.
The files are linearly dependent due to limitations of Abella.

\begin{figure}[t]
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  311 & 57  & Basic data structures\\\hline
    typing.thm          &  245 & 7   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strengthening\\\hline
    alg.thm             &  679 & 80  & Basic algorithmic properties\\\hline
    trans.thm           &  616 & 53  & \makecell[l]{Worklist instantiation and declarative transfer\\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  909 & 70  & \makecell[l]{Non-overlapping declarative system \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll},
                                \ref{lem:inv_chkLam}, \ref{lem:subsumption}}\\\hline
    soundness.thm       & 1107 & 78  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & The $|\cdot|_\to$ measure; Lemma~\ref{lem:prune_inst}\\\hline
    dcl.thm             &  380 & 12  & Non-overlapping declarative worklist \\\hline
    completeness.thm    & 1124 & 61  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  837 & 45  & Other worklist measures; Lemma~\ref{lemma:inst:decidable}\\\hline
    decidability.thm    &  983 & 57  & Decidability theorem and corollary\\\hline
    smallStep.thm       &  119 &  2  & The equivalence between big-step and small-step\\\hline
    \hline\emph{Total}               & 7977 & 596 & (60 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}

### 2019-05-30 sections/Overview.tex
shrinking text

\centering \framebox{$\Psi \vdash A \le B$}
\begin{gather*}
\inferrule*[right=$\mathtt{{\le}Var}$]
    {a\in\Psi}{\Psi\vdash a\le a}
\qquad
\inferrule*[right=$\mathtt{{\le}Unit}$]
    {~}{\Psi \vdash 1 \le 1}
\qquad
\inferrule*[right=$\mathtt{{\le}{\to}}$]
    {\Psi \vdash B_1 \le A_1 \quad \Psi \vdash A_2 \le B_2}
    {\Psi\vdash A_1\to A_2 \le B_1\to B_2}
\\
\inferrule*[right=$\mathtt{{\le}\forall L}$]
    {\Psi\vdash \tau \quad \Psi\vdash [\tau/a] A \le B}
    {\Psi\vdash \all A \le B}
\qquad
\inferrule*[right=$\mathtt{{\le}\forall R}$]
    {\Psi, b\vdash A\le B}
    {\Psi\vdash A \le \all[b]B}
\end{gather*}
\caption{%Well-formedness of Declarative Types and 
Declarative Subtyping}\label{fig:decl:sub}
\end{figure}

@@

\jimmy{Mention the implicit freshness conditions in the premises}
Please note that when we introduce a new binder in the premise, we implicitly pick a fresh one.
This applies to rules such as $\mathtt{wf_d\forall}$, $\mathtt{wf_dabs}$, $\mathtt{{\le}\forall R}$,
throughout the whole text.

@@

%-------------------------------------------------------------------------------
\subsection{Judgment Lists}\label{sec:overview:list}
To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a number
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted everywhere.

@@

The algorithm first analyses the two subtyping judgments and
will find the best solutions $\al = \bt = 1$
(please refer to Figure~\ref{fig:alg} for detailed derivations).
Then we substitute every instance of $\al$ and $\bt$ with $1$,
so the variable declarations can be safely removed from the worklist.
Now we reduce the worklist to the following state
$$\nil \Vdash \underline{(\lam x)~() \To_a a \le 1}, x:1$$
which has a term variable declaration as the top element.
%Because no judgments before the declaration refer to $x$,
%as a basic property of an ordered context
%(Figure~\ref{fig:alg:syntax} shows its well-formedness criteria),
%we remove all declarations before the first judgment chain to collect unused variables.
%As a result all garbage variables are collected
%once the judgment chains refering to them are processed,
%just as what we have mentioned above about the scope tracking.
After removing the garbage term variable declaration from the worklist, we process the last remaining judgment chain,
namely an inference judgment $(\lam x)~()\To~?$, with the unit type $1$ as its result.
Finally the last judgment becomes $1 \le 1$, a trivial base case.

>>>>

\centering \framebox{$\Psi \vdash A \le B$}
\begin{gather*}
\inferrule*[right=$\mathtt{{\le}Var}$]
    {a\in\Psi}{\Psi\vdash a\le a}
\qquad
\inferrule*[right=$\mathtt{{\le}Unit}$]
    {~}{\Psi \vdash 1 \le 1}
\qquad
\inferrule*[right=$\mathtt{{\le}{\to}}$]
    {\Psi \vdash B_1 \le A_1 \quad \Psi \vdash A_2 \le B_2}
    {\Psi\vdash A_1\to A_2 \le B_1\to B_2}
\\
\inferrule*[right=$\mathtt{{\le}\forall L}$]
    {\Psi\vdash \tau \quad \Psi\vdash [\tau/a] A \le B}
    {\Psi\vdash \all A \le B}
\qquad
\inferrule*[right=$\mathtt{{\le}\forall R}$]
    {\Psi, b\vdash A\le B}
    {\Psi\vdash A \le \all[b]B}
\end{gather*}
\caption{%Well-formedness of Declarative Types and 
Declarative Well-formedness and Subtyping}\label{fig:decl:sub}
\end{figure}

@@

% \jimmy{Mention the implicit freshness conditions in the premises}
Please note that when we introduce a new binder in the premise, we implicitly pick a fresh one.
This applies to rules such as $\mathtt{wf_d\forall}$, $\mathtt{wf_dabs}$, $\mathtt{{\le}\forall R}$,
throughout the whole text.

@@

%-------------------------------------------------------------------------------
\subsection{Judgment Lists}\label{sec:overview:list}
To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a number
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted.

@@

The algorithm first analyses the two subtyping judgments and
will find the best solutions $\al = \bt = 1$
(please refer to Figure~\ref{fig:alg} for detailed derivations).
Then we substitute every instance of $\al$ and $\bt$ with $1$,
so the variable declarations can be safely removed from the worklist.
Now we reduce the worklist to the following state
$$\nil \Vdash \underline{(\lam x)~() \To_a a \le 1}, x:1$$
which has a term variable declaration as the top element.
%Because no judgments before the declaration refer to $x$,
%as a basic property of an ordered context
%(Figure~\ref{fig:alg:syntax} shows its well-formedness criteria),
%we remove all declarations before the first judgment chain to collect unused variables.
%As a result all garbage variables are collected
%once the judgment chains refering to them are processed,
%just as what we have mentioned above about the scope tracking.
After removing the garbage term variable declaration from the worklist, we process the last remaining
inference judgment $(\lam x)~()\To~?$, with the unit type $1$ as its result.
Finally the last judgment becomes $1 \le 1$, a trivial base case.

### 2019-05-30 sections/Related.tex
shrinking text

Algorithm $\mathcal{W}$,
as one of the classic type inference algorithms for polymorphic type systems,
has been manually proven to be sound and complete
with respect to the Hindley-Milner type system~\cite{hindley1969principal,milner1978theory,damas1982principal}.
After around 15 years, the algorithm was formally verified by
\citet{naraschewski1999type} in Isabelle/HOL~\cite{nipkow2002isabelle}.
The treatment of new variables was tricky at that time, while the overall structure follows the
structure of Damas's manual proof closely.
%Another complication is the encoding of substitutions,
%which they chose to formalize as a function instead of an association list.
Later on, other researchers~\cite{dubois2000proving,dubois1999certification}
formalized algorithm $\mathcal{W}$ in Coq~\cite{Coq}.
Nominal techniques~\cite{urban2008nominalTech} in Isabelle/HOL have been
developed to help programming language formalizations, and are used for a similar
verification~\cite{urban2008nominal}. Moreover, Garrigue~\cite{garrigue2015certified}
mechanized a type inference algorithm,
with the help of locally nameless cofinite quantification~\cite{LocallyNameless},
for Core ML extended with structural polymorphism and recursion.

\paragraph{Ordered Contexts in Type Inference}
Gundry et al.~\cite{gundry2010type} revisit algorithm $\mathcal{W}$ and
propose a new unification algorithm with the help of ordered contexts.
Similar to DK's algorithm, information of meta variables flow from input contexts to output contexts.
Not surprisingly, its information increase relation has a similar role to DK's context extension.
Our algorithm, in contrast,
eliminates the use of output contexts and their variable solution records ($\al = \tau$),
simplifying the information propagation process through immediate substitution
by collecting all the judgments in a single worklist.

@@

\paragraph{Lists of Judgments in Unification}
Some work~\cite{Reed2009,Abel2011higher} adopts a similar idea to this paper
in work on unification for dependently typed languages. Similarly to our work
the algorithms need to be very careful about scoping, since the order of variable
declarations is fundamental in a dependently typed setting. 
Their algorithms simplify a collection of unification constraints progressively in a single-step style.
In comparison, our algorithm mixes variable declarations with judgments,
resulting in a simpler judgment form,
while processing them in a similar way to those algorithms.
One important difference is that contexts are
duplicated in their unification judgments, which complicates the unification process,
since the information of each local context needs to be synchronized.
Instead we make use of the nature of ordered context to control the scoping of unification variables.
While their algorithms focus only on unification,
our algorithm also deals with other types of judgments like type inference.
A detailed discussion is in Section~\ref{sec:overview:list}.

>>>>

Algorithm $\mathcal{W}$,
as one of the classic type inference algorithms for polymorphic type systems,
has been manually proven to be sound and complete
with respect to the Hindley-Milner type system~\cite{hindley1969principal,milner1978theory,damas1982principal}.
After around 15 years, the algorithm was formally verified by
\citet{naraschewski1999type} in Isabelle/HOL~\cite{nipkow2002isabelle}.
The treatment of new variables was tricky at that time, while the overall structure follows the
structure of Damas's manual proof closely.
%Another complication is the encoding of substitutions,
%which they chose to formalize as a function instead of an association list.
Later on, other researchers~\cite{dubois2000proving,dubois1999certification}
formalized algorithm $\mathcal{W}$ in Coq~\cite{Coq}.
Nominal techniques~\cite{urban2008nominalTech} in Isabelle/HOL have been
developed to help programming language formalizations, and are used for a similar
verification~\cite{urban2008nominal}. Moreover, Garrigue~\cite{garrigue2015certified}
mechanized a type inference algorithm,
with the help of locally nameless~\cite{LocallyNameless},
for Core ML extended with structural polymorphism and recursion.

\paragraph{Ordered Contexts in Type Inference}
Gundry et al.~\cite{gundry2010type} revisit algorithm $\mathcal{W}$ and
propose a new unification algorithm with the help of ordered contexts.
Similar to DK's algorithm, information of meta variables flow from input contexts to output contexts.
Not surprisingly, its information increase relation has a similar role to DK's context extension.
Our algorithm, in contrast,
eliminates output contexts and solution records ($\al = \tau$),
simplifying the information propagation process through immediate substitution
by collecting all the judgments in a single worklist.

@@

\paragraph{Lists of Judgments in Unification}
Some work~\cite{Reed2009,Abel2011higher} adopts a similar idea to this paper
in work on unification for dependently typed languages. Similarly to our work
the algorithms need to be very careful about scoping, since the order of variable
declarations is fundamental in a dependently typed setting. 
Their algorithms simplify a collection of unification constraints progressively in a single-step style.
In comparison, our algorithm mixes variable declarations with judgments,
resulting in a simpler judgment form,
while processing them in a similar way.
One important difference is that contexts are
duplicated in their unification judgments, which complicates the unification process,
since the information of each local context needs to be synchronized.
Instead we make use of the nature of ordered context to control the scopes of unification variables.
While their algorithms focus only on unification,
our algorithm also deals with other types of judgments like synthesis.
A detailed discussion is in Section~\ref{sec:overview:list}.

### 2019-05-30 sections/algorithm.tex
shrinking text

Figure~\ref{fig:alg:syntax} shows the syntax and well-formedness
judgments used by the algorithm. Similarly to the declarative system 
the well-formedness rules are unsurprising and merely ensure 
well-scoping of variables.

@@

\jimmy{The implicit freshness conditions for algorithmic system}

Please note that when a new variable is introduced in the right-hand side worklist $\Gm'$,
we implicitly pick a fresh one,
since the right-hand side can be seen as the premise of the reduction.

@@

\paragraph{Example} The derivation of the subtyping judgment
$(1\to 1)\to 1 \le (\all 1\to 1) \to 1$ proceeds as follows:
\begin{gather*}
\begin{aligned}
  & \nil\vdash (1\to 1)\to 1 \le (\all 1 \to 1) \to 1\\
\rrule{7}  & \nil \Vdash 1 \le 1 \Vdash \all 1 \to 1 \le 1 \to 1\\
\rrule{8}  & \nil \Vdash 1 \le 1 ,\al \Vdash 1 \to 1 \le 1 \to 1\\
\rrule{7}  & \nil \Vdash 1 \le 1 ,\al \Vdash 1 \le 1 \Vdash 1 \le 1\\
\rrule{4} & \nil \Vdash 1 \le 1 ,\al \Vdash 1 \le 1\\
\rrule{4} & \nil \Vdash 1 \le 1 ,\al\\
\rrule{2} & \nil \Vdash 1 \le 1\\
\rrule{4} & \nil
\end{aligned}
\end{gather*}
First, the subtyping of two function types is split into two judgments by Rule 7:
a covariant subtyping on the return type and a contravariant subtyping on the argument type.
Then we apply Rule 8 to reduce the $\forall$ quantifier on the left side.
The rule introduces an existential variable $\al$ to the context (even though 
the type $\all 1 \to 1$ does not actually refer to the quantified type
variable $a$).
In the following 3 steps we satisfy the judgment $1 \to 1 \le 1 \to 1$ by Rules 7, 4 and 4.

@@

The alternative would have been to omit Rule 21 and drop the condition on $e$
in Rule 18. The modified Rule 18 would then handle $\Gm \Vdash \lam e \Lto \al$
and yield $\Gm \Vdash \lam e \To_a a \le \al$, which would be further processed
by Rule 25 to $\Gm, \bt[1], \bt[2] \Vdash \bt[1] \to \bt[2] \le \al, x:\bt[1] \Vdash e \Lto \bt[2]$.
As a subtyping constraint between two monotypes only succeeds when the types are equal, $\bt[1] \to \bt[2] \le \al$
must end up equating $\bt[1] \to \bt[2]$ with $\al$ and thus have the same effect as Rule 21, but in a more roundabout
fashion.

>>>>

Figure~\ref{fig:alg:syntax} shows the syntax and well-formedness
judgments used by the algorithm. Similarly to the declarative system 
the well-formedness rules are unsurprising and merely ensure 
well-scopedness.

@@

% \jimmy{The implicit freshness conditions for algorithmic system}
Please note that when a new variable is introduced in the right-hand side worklist $\Gm'$,
we implicitly pick a fresh one,
since the right-hand side can be seen as the premise of the reduction.

@@

\paragraph{Example} The derivation of the subtyping judgment
$(1\to 1)\to 1 \le (\all 1\to 1) \to 1$:
\begin{gather*}
\begin{aligned}
  & \nil\vdash (1\to 1)\to 1 \le (\all 1 \to 1) \to 1\\
\rrule{7}  & \nil \Vdash 1 \le 1 \Vdash \all 1 \to 1 \le 1 \to 1\\
\rrule{8}  & \nil \Vdash 1 \le 1 ,\al \Vdash 1 \to 1 \le 1 \to 1\\
\rrule{7}  & \nil \Vdash 1 \le 1 ,\al \Vdash 1 \le 1 \Vdash 1 \le 1\\
\rrule{4} & \nil \Vdash 1 \le 1 ,\al \Vdash 1 \le 1\\
\rrule{4} & \nil \Vdash 1 \le 1 ,\al\\
\rrule{2} & \nil \Vdash 1 \le 1\\
\rrule{4} & \nil
\end{aligned}
\end{gather*}
First, the subtyping of two function types is split into two judgments by Rule 7:
a covariant subtyping on the return type and a contravariant subtyping on the argument type.
Then we apply Rule 8 to reduce the $\forall$ quantifier on the left side.
The rule introduces an existential variable $\al$ to the context (even though 
the type $\all 1 \to 1$ does not actually refer to the quantified type
variable $a$).
In the following 3 steps we satisfy the judgment $1 \to 1 \le 1 \to 1$ by Rules 7, 4 and 4.

@@

The alternative would have been to omit Rule 21 and drop the condition on $e$
in Rule 18. The modified Rule 18 would then handle $\Gm \Vdash \lam e \Lto \al$
and yield $\Gm \Vdash \lam e \To_a a \le \al$, which would be further processed
by Rule 25 to $\Gm, \bt[1], \bt[2] \Vdash \bt[1] \to \bt[2] \le \al, x:\bt[1] \Vdash e \Lto \bt[2]$.
As a subtyping constraint between monotypes is simply equality, $\bt[1] \to \bt[2] \le \al$
must end up equating $\bt[1] \to \bt[2]$ with $\al$ and thus have the same effect as Rule 21, but in a more roundabout
fashion.
